---
title: "Introduction: pre-post and diff-in-diff"
subtitle: "Causal impact assessment workshop"
author: "Erik-Jan van Kesteren & Oisín Ryan"
eval: false
format: 
  html:
    toc: true
    self-contained: true
    code-fold: true
    df-print: paged
    callout-appearance: simple
    callout-icon: false
---

This is the first practical, where we introduce the dataset and we use it to create pre-post and diff-in-diff estimates for the causal effect of the California proposition 99 policy intervention.

You can use your preferred way of working in R to do the practicals. Our preferred way is this:

- Create a new folder with a good name, e.g., `practicals_causal_impact` 
- Open RStudio
- Create a new project from RStudio, which you associate with the folder
- Create a `raw_data` subfolder
- Create an R script for the current practical, e.g., `introduction.R`
- Create your well-documented and [well-styled](https://style.tidyverse.org/) code in this R script


::: {.callout-tip appearance="default" icon=true}
The answers to each exercise are available as a collapsed `code` block. Try to work out the answer yourself before looking at this code block!
:::


In all practicals in this workshop, we make extensive use of the `tidyverse` set of packages. You can load these packages like so:

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-fold: false
#| eval: true
library(tidyverse)
```

In this practical, we will also use the following two packages:

```{r}
#| label: setup-2
#| code-fold: false
library(sandwich)
library(lmtest)
```
## The data
We will be using the `proposition99` dataset that we introduced in the lecture. We have prepared the dataset for you to download [`here`](https://github.com/sodascience/workshop_causal_impact_assessment/raw/main/data/proposition99.rds). It is an `rds` file, which is a convenient, portable, and fast binary file format for R.

::: {.callout-note}
## Exercise 1
Download the dataset and save it in a nice location, e.g., a `raw_data` folder inside your R project.
:::


::: {.callout-note}
## Exercise 2
Load the dataset in R using the `tidyverse` function `read_rds()`. Give the dataset the name `prop99`. Then, inspect the first few rows of the data.

```{r}
#| label: data-true
#| echo: false
#| eval: true
prop99 <- read_rds("../../data/proposition99.rds")
```

```{r}
#| label: data

# read the dataset to a variable called prop99
prop99 <- read_rds("raw_data/data/proposition99.rds")

# inspect the first few rows
head(prop99)
```
:::


::: {.callout-note}
## Exercise 3
Using `filter()`. `group_by()`, `summarize()`, and `arrange()`, find out which state had the highest average retail price of a box of cigarettes before 1988.

```{r}
#| label: retailprice

# read the dataset to a variable called prop99
prop99 |> 
  filter(year < 1988) |> 
  group_by(state) |> 
  summarize(price = mean(retprice)) |> 
  arrange(desc(price))

```
:::

## Pre-post estimator
In this section, you will estimate the causal effect of the policy using the pre-post estimator. For this, you need to select only California from the data, then create a factor variable for the pre and post period, and then use linear regression to estimate the causal effect.

::: {.callout-note}
## Exercise 4
Use `filter()` to select only California from the dataset and use `mutate()` to create a pre-post indicator variable called `prepost`. Remember: include the year 1988 in the pre-period. Make sure your `prepost` variable is of the type `factor`. Assign the result to a variable called `prop99_cali`.

```{r}
#| label: prepostprep

# create the pre-post dataset
prop99_cali <- 
  prop99 |> 
  filter(state == "California") |> 
  mutate(prepost = as_factor(ifelse(year <= 1988, "Pre", "Post"))) 
```
:::

In the lecture, we chose to include 12 years before and after the intervention. In this practical, we will use only 5 years before and after the intervention for our effect estimate.

::: {.callout-note}
## Exercise 5
Use `filter()` to include data between 1984 and 1993. Then, use linear regression (`lm()`) to estimate the effect of the proposition 99 intervention, then use `summary` on the fitted model object to look at the estimate. Is this effect different from the one estimated in the lecture?

```{r}
#| label: prepostest

# fit the model with 5 years pre and post
fit_prepost <- lm(
  formula = cigsale ~ prepost, 
  data = prop99_cali |> filter(year > 1983, year < 1994)
)

# investigate the effect
summary(fit_prepost)

# the effect estimated in this way is -27.020
# this is much smaller than in the lecture!
```
:::


In the lecture, we did not correct the inference (p-value) for potential autocorrelation. We can do this with the function `coeftest()` on our fitted model object.

::: {.callout-note}
## Exercise 6
Use `coeftest()` to correct the inference using heteroscedasticity and autocorrelation consistent (HAC) standard errors (pass the `vcovHAC` function from the `sandwich` package to the `.vcov` argument). Is the pre-post causal effect significantly different from 0?

```{r}
#| label: HAC-correct
coeftest(fit_prepost, vcov. = vcovHAC)

# The standard error is a little bigger
# (it is now 5.29 versus 4.34 before)
# but the effect is still significant at 
# the 5% level. (p < .001)
```
:::

## Difference-in-differences estimator

In this section, we select a suitable control state to perform a diff-in-diff estimate of the causal effect of the policy intervention. In this section, you will not choose Utah as a control state as in the lectures, but one of the following states:

- Nevada
- Montana
- Colorado

Here are the data plots for these three states:

```{r}
#| label: didplots
#| eval: true

# Diff-in-diff time series figure
prop99 |> 
  filter(state %in% c("California", "Nevada", "Montana", "Colorado")) |> 
  ggplot(aes(x = year, y = cigsale, colour = state)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 1988, lty = 2) +
  theme_minimal() +
  scale_colour_manual(values = c("orange", "#AA8888",  "#88AA88","#8888AA")) +
  annotate("label", x = 1988, y = 150, label = "Intervention") +
  labs(title = "Panel data for California three potential control states",
       y = "Cigarette sales", x = "Year", colour = "")
```


::: {.callout-note}
## Exercise 7
Create a dataset called `prop99_did` which includes California and your chosen control state. As before, create a `prepost` variable and include only the 5 years before and after the intervention.

```{r}
#| label: didprep

# prepare the did data
prop99_did <- 
  prop99 |> 
  filter(
    state == "California" | state == "Nevada", 
    year > 1983, year < 1994
  ) |> 
  mutate(prepost = as_factor(ifelse(year <= 1988, "Pre", "Post"))) |> 
  filter()
```
:::


::: {.callout-note}
## Exercise 8
Now, estimate the causal effect using the difference-in-differences estimator. For this, use the formula `cigsale ~ state * prepost` in the `lm()` function. Investigate the estimated effect using HAC standard errors as before. How big is the causal effect of the policy intervention and is this effect significantly different from 0?
```{r}
#| label: didestimate

# fit the model with 5 years pre and post
fit_did <- lm(
  formula = cigsale ~ state * prepost, 
  data = prop99_did
)

# investigate the effect
coeftest(fit_did, vcov. = vcovHAC)

# Using Nevada as a control,
# the did causal effect is 5.68,
# with a HAC s.e. of 5.42
# so this effect is not significantly
# different from 0.
```
:::

## Conclusion

You have created causal effect estimates using a pre-post design and using a diff-in-diff design, and you have corrected the inferences using heteroskedasticity and autocorrelation consistent standard errors. You have seen that the conclusions are very dependent on the choices made, for example about which period to consider and which control unit to choose.
