---
title: "Synthetic control: tidysynth, inference, robustness"
subtitle: "Causal impact assessment workshop"
author: "Erik-Jan van Kesteren & Oisín Ryan"
eval: false
format: 
  html:
    toc: true
    self-contained: true
    code-fold: true
    df-print: paged
    callout-appearance: simple
    callout-icon: false
---

In this practical, we will use the following two packages:

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-fold: false
#| eval: true
library(tidyverse)
library(tidysynth)
```

We will again be using the proposition 99 dataset:
```{r}
#| label: data
#| code-fold: false

prop99 <- read_rds("raw_data/proposition99.rds")
```

```{r}
#| label: data-true
#| echo: false
#| eval: true
prop99 <- read_rds("../../data/proposition99.rds")
```

In the following sections, you will create an causal effect estimate using the synthetic control method, you will perform inference for this estimate, and you will do some robustness checks.

## Synthetic control in tidysynth

The first step in the `tidysynth` package framework is to create an object from the dataset that will provide the basis of the estimation method.

::: {.callout-note}
## Exercise 1
Use the the `synthetic_control()` function to create a synthetic control object from the `prop99` data. Read the help file to (`?synthetic_control`) if you need to know more about the arguments needed. Set the argument `generate_placebos = TRUE` (we will need this later).

```{r}
#| label: synthcontrol-object

# create a synthetic control object
prop99_syn <- 
  prop99 |> 
  synthetic_control(
    outcome = cigsale,
    unit = state,
    time = year,
    i_unit = "California",
    i_time = 1988,
    generate_placebos = TRUE
  )

```
:::

In tidysynth, the `grab_*()` functions can be used to inspect the object in detail. For example, you can use `grab_outcome()` we inspect the outcome (cigsale) for the treated unit and the potential controls.

::: {.callout-note}
## Exercise 2
Inspect the outcome variable for the treated and the control units, to check that everything worked as you expected.

```{r}
#| label: inspect-outcome

# Treated unit (california)
grab_outcome(prop99_syn)

# control units
grab_outcome(prop99_syn, type = "controls")
```
:::

The next step is to determine and create the variables that will be used for matching and estimating weights. These are called "predictors" in `tidysynth`. 


::: {.callout-note}
## Exercise 3
Generate the following predictors using the `generate_predictor()` function. You will need to run this function multiple times, once for each time period considered.

- Mean (log-)income in 1980-1988
- Mean retail price of cigarettes in 1980-1988
- Mean proportion of people aged 15 to 24 in 1980-1988
- Mean beer consumption in 1984-1988
- Cigarette sales in 1975
- Cigarette sales in 1980
- Cigarette sales in 1988

NB: there are some missing values in this data, so use `na.rm = TRUE` inside your `mean()` function.

```{r}
#| label: predictors

# create predictors
prop99_syn <- 
  prop99_syn |> 
  generate_predictor(
    time_window = 1980:1988,
    lnincome = mean(lnincome, na.rm = TRUE),
    retprice = mean(retprice, na.rm = TRUE),
    age15to24 = mean(age15to24, na.rm = TRUE)
  ) |> 
  generate_predictor(
    time_window = 1984:1988,
    beer = mean(beer, na.rm = TRUE)
  ) |> 
  generate_predictor(
    time_window = 1975,
    cigsale_1975 = cigsale
  ) |> 
  generate_predictor(
    time_window = 1980,
    cigsale_1980 = cigsale
  ) |> 
  generate_predictor(
    time_window = 1988,
    cigsale_1988 = cigsale
  )
```
:::


## Inference using permutation test

section 2

## Robustness checks for units and variables

section 3

## Conclusion

Section 4