---
title: "Synthetic control: tidysynth, inference, robustness"
subtitle: "Causal impact assessment workshop"
author: "Erik-Jan van Kesteren & Oisín Ryan"
eval: false
format: 
  html:
    toc: true
    self-contained: true
    code-fold: true
    df-print: paged
    callout-appearance: simple
    callout-icon: false
---

In this practical, we will use the following two packages:

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-fold: false
#| eval: true
library(tidyverse)
library(tidysynth)
```

We will again be using the proposition 99 dataset that we prepared.

::: {.callout-note}
## Exercise 1
Load the `proposition99` dataset in R using the `tidyverse` function `read_rds()`. Give the dataset the name `prop99`. 

```{r}
#| label: data-true
#| echo: false
#| eval: true
prop99 <- read_rds("../../data/proposition99.rds")
```

```{r}
#| label: data

# read the dataset to a variable called prop99
prop99 <- read_rds("raw_data/data/proposition99.rds")
```
:::

## Synthetic control in tidysynth

section 1

## Inference using permutation test

section 2

## Robustness checks for units and variables

section 3

## Conclusion

Section 4